---
title: "R Notebook"
output: 
  html_notebook: 
    toc: yes
---

# Week 1
## Matrix Derivatives

If we want to explain a vector $y_{n \times 1}$ with $X_{n \times p}$, we need a vector $\beta_{p \times 1}$ that minimize $$\lVert y_{n \times 1} - X_{n \times p}\beta_{p \times 1} \rVert$$
And the minimum of this expression is $$ \hat{\beta} = (x^tx)^{-1}x^ty $$

### Example
```{r}
data(mtcars)
head(mtcars)
```
We define our y, x and $\beta$ variables

\%\*\% matrix multiplication

solve(a) calculates inverse of a

solve(a,b) solves equation a %*% x = b

```{r}
y = mtcars$mpg
x = cbind(1,mtcars$wt, mtcars$hp)
beta = solve(t(x) %*% x) %*% t(x) %*% y
beta
```
We can check that we obtain the same vector than if we use the function lm
```{r}
coef(lm(mpg ~ wt + hp, data = mtcars))
```
## Centering by matrix multiplication
Let $J_n$ be a vector full of 1 of length n. Then, the matrix $$ (I - J_n(J_n^tJ_n)^{-1}J_n^t) = (I - \frac{1}{n}J_{n \times n}) $$ with $J_{n \times n}$ a matrix full of 1, multiplying $$ (I - \frac{1}{n}J_{n \times n})X_{n \times p} $$ will mean-center the columns of X

### Example

```{r}
y = mtcars$mpg
x = cbind(1,mtcars$wt, mtcars$hp)
n = nrow(x)
I = diag(rep(1,n))
H = matrix(1,n,n)/n
xc = (I-H) %*% x
```

If we calculate the mean by columns then we get zeros

```{r}
apply(xc,2,mean)
```
Same thing the other way around

```{r}
y = mtcars$mpg
x = cbind(1,mtcars$wt, mtcars$hp)
p = 3
I = diag(rep(1,p))
H = matrix(1,p,p)/p
xc = x %*%(I-H) 
apply(xc,1,mean)
```
 More efficient way using function sweep
 
```{r}
xc2 = sweep(x,2,apply(x,2,mean))
apply(xc2,2,mean)
```
 
 